---
title: "SpatialAnalysis"
author: "Maggie O'Shea"
date: "4/12/2022"
output: html_document
---
#Maggie note to self (where left off)
* Merged files but data won't transfer from flood properties to spatial data - waiting to hear from John --> should be easy to map from there 

```{r setup, include=FALSE}
library(tidyverse, quietly = TRUE)
library(lubridate)

#install.packages('sf')
library(sf)
#install.packages('leaflet')
library(leaflet)
#install.packages('mapview')
library(mapview)
library(priceR)

raw_flooddata <- read.csv("./Data/Raw_Data/HazardMitigationAssistanceMitigatedProperties.csv", stringsAsFactors = TRUE)

inflationadjusted <- read.csv('./Data/inflationadjusted.csv')
#Removed shapefile from code because it was too big to push up 
#USA_counties <- st_read('./Data/Raw_Data/USA_Counties/USA_Counties.shp')
NC_countyboundaries <- st_read('./Data/Raw_Data/North_Carolina_State_and_County_Boundary_Polygons')

```


## Wrangle by County

# Actual Amount Paid 
```{r}
#All flood properties
NC_clean_hazardprops <- raw_flooddata %>%
  filter(!is.na(actualAmountPaid))%>%
  filter(actualAmountPaid > 0)%>%
  filter(state == "North Carolina")

NC.floodproperties_amountpaid <- NC_clean_hazardprops %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10|202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(county, state)%>%
  summarise(All_propertiesamountpaid = sum(actualAmountPaid))

#Ex situ
NC.buyoutfloodproperties_amountpaid <- NC_clean_hazardprops %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10', type))%>%
  group_by(county, state)%>%
  summarise(Buyouts_propertiesamountpaid = sum(actualAmountPaid))

#In situ
NC.insitufloodproperties_amountpaid <- NC_clean_hazardprops %>%
  filter(grepl('202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(county, state)%>%
  summarise(insituproperties_amountpaid = sum(actualAmountPaid))

in.ex.county <- left_join(NC.buyoutfloodproperties_amountpaid, NC.insitufloodproperties_amountpaid, by = c("county", "state"))

amountpaid.NCall <- left_join(in.ex.county, NC.floodproperties_amountpaid, by = c("county", "state"))

```

# Amount Paid Inflation Adjustments

```{r}
cleaner_hazardprops <- NC_clean_hazardprops%>%
  filter(county != "")

wYear_NC.paid <- cleaner_hazardprops  %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10|202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(county, state, programFy)%>%
  summarise(All_propertiesamountpaid = sum(actualAmountPaid))

#Ex situ
wyear_NC.buyoutpaid<- cleaner_hazardprops  %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10', type))%>%
  group_by(county, state, programFy)%>%
  summarise(Buyouts_propertiesamountpaid = sum(actualAmountPaid))

#In situ
wyear_NC.insitupaid <- cleaner_hazardprops  %>%
  filter(grepl('202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(county, state, programFy)%>%
  summarise(insituproperties_amountpaid = sum(actualAmountPaid))


Allpaid_inflation <- priceR::adjust_for_inflation(wYear_NC.paid$All_propertiesamountpaid, wYear_NC.paid$programFy, "US", to_date = 2020)
vector_allinflation <- c(Allpaid_inflation)

inflat_paidAll <- wYear_NC.paid %>%
  mutate('Allpaid_inflation' = vector_allinflation)

#Ex situ

exsitupaid_inflation <- priceR::adjust_for_inflation(wyear_NC.buyoutpaid$Buyouts_propertiesamountpaid, wyear_NC.buyoutpaid$programFy, "US", to_date = 2020)

inflation_buyoutpaid.county <- wyear_NC.buyoutpaid%>%
  mutate("InflationAdjusted_BUYOUTAmountpaid" = exsitupaid_inflation)

#In situ
insitupaid_inflation <- priceR::adjust_for_inflation(wyear_NC.insitupaid$insituproperties_amountpaid, wyear_NC.insitupaid$programFy, "US", to_date = 2020)

inflation_insituprops_amountpaid <- wyear_NC.insitupaid%>%
  mutate("InflationAdjusted_INSITUAmountpaid" = insitu_inflation)
```

# Number of Properties

```{r}
NC.floodproperties_number <- NC_clean_hazardprops %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10|202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(county, state)%>%
  summarise(All_properties = sum(numberOfProperties))

#Ex situ
NC.buyoutfloodproperties <- NC_clean_hazardprops %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10', type))%>%
  group_by(county, state)%>%
  summarise(Buyouts_propertiesperyear = sum(numberOfProperties))

#In situ
NC.insitufloodproperties <- NC_clean_hazardprops  %>%
  filter(grepl('202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(county, state)%>%
  summarise(insitupropertiesperyear = sum(numberOfProperties))

insitu.props <- left_join(NC.insitufloodproperties, NC.buyoutfloodproperties, by = c("county", "state"))
numberofprops.NC <- left_join(insitu.props, NC.floodproperties_number, by = c("county", "state"))
```

```{r make and wrangle combiined dataset}
alldata.bycountyNC <- left_join(numberofprops.NC, amountpaid.NCall, by = c("county", "state"))%>%
  mutate("County" = county)

#Turn NAs to zeros because [assumption we are making is..] those counties with no value in column indicates that they did not receive a grant

alldata.bycountyNC[is.na(alldata.bycountyNC)] <- 0  

flood_counties_join <- left_join(NC_countyboundaries, alldata.bycountyNC, by = "County")
flood_counties_merge <-  merge(x = NC_countyboundaries,
                           y = alldata.bycountyNC, 
                           by.x = 'County', 
                           by.y = 'county')

#Turn NAs to zeros again for added counties from map 
flood_counties_merge [is.na(flood_counties_merge)] <- 0  
flood_counties_join [is.na(flood_counties_join)] <- 0 

class(flood_counties_join)
```


```{r}
mapview(flood_counties_join, zcol='All_properties')

library(RColorBrewer)
library(viridis)

#paletteer::scale_fill_paletteer_d("wesanderson::FantasticFox1")
  #scale_fill_continuous(type = 'viridis', direction = -1)
    #  scale_fill_gradient2(
    #low = "cadetblue1", 
  # high = "deepskyblue2" )

ggplot(data=flood_counties_join) +
  geom_sf(aes(fill=All_properties)) + 
  scale_fill_gradientn(colours = colorspace::heat_hcl(7), trans = 'reverse')+
  labs(
    title='Number of Properties that Received Flooding Adaptation Funding: 1985-2020',
    fill ='Number of Properties'
  )+
  theme(plot.title = element_text(size=12, hjust=0.5), 
        legend.text = element_text(size = 10))
```


```{r}
ggplot(data=flood_counties_join) +
  geom_sf(aes(fill=insitupropertiesperyear)) + 
  scale_fill_gradientn(colours = colorspace::heat_hcl(7), trans = 'reverse')+
  labs(
    title='Number of Properties that Received In Situ Adaptation Funding: 1985-2020',
    fill ='Number of Properties'
  )+
  theme(plot.title = element_text(size=12, hjust=0.5), 
        legend.text = element_text(size = 10))

ggplot(data=flood_counties_join) +
  geom_sf(aes(fill=Buyouts_propertiesperyear)) + 
  scale_fill_gradientn(colours = colorspace::heat_hcl(7), trans = 'reverse')+
  labs(
    title='Number of Properties that Received Ex Situ Adaptation Funding: 1985-2020',
    fill ='Number of Properties'
  )+
  theme(plot.title = element_text(size=12, hjust=0.5), 
        legend.text = element_text(size = 10))
```
# Notes from Maggie: 
This is interesting because it shows that buyouts are concentrated near what is likely riverine flooding, while coastal flooding gets more in-situ adaptation money. You can also see in the coastal region that inland counties have more buyouts than coastal which could indicate that wealthier coastal homeowners are able to do more to keep their home where it is, whereas inland communities are being bought out. 


