---
title: "Wrangling data"
author: "Maggie O'Shea"
date: "4/2/2022"
output: html_document
editor_options: 
  chunk_output_type: inline
---
###Key to Dataset Names: 
Note: There are other datasets that I made in wrangling, but it was simply to get to the final outcome datasets, especially when I was merging sets. Those in-between ones aren't listed here

**hazard_properties** = Raw Hazard Property Data
**clean_hazardprops** = Hazard Property Data with the negative numbers removed 
**SOVI_USA** = Social Vulnerability Dataset from ESRI
**SOVI_wrangle** = SVI data with added percent owner occupied column

#Time Series Relevant Data: 
--------------------------
**floodproperties_amountpaid** = Total USD Paid *per year* for all properties funded to adapt to flooding

**buyoutfloodproperties_amountpaid** = Total USD paid *per year* for properties bought out due to flooding =
**insitufloodproperties_amountpaid** = Total USD paid *per year* for properties to adapt in place (in situ), aka for owners to stay where they are and re-construct their home/property to be more resilient to floods

**floodproperties** = Number of properties that received grant money for flooding hazards *per year*

**buyoutfloodproperties** =  Number of properties that were bought out due to flooding *per year*
**insitufloodproperties** = Number of properties that received grant money to adapt the property in situ *per year* 

**propertiesvalue_allflood** = Joined number of properties and amount paid *per year* 

**zero_NA_alldata** = Joined all *per year* data including all, in-situ, and ex-situ for number of properties and total USD paid with NAs turned to zeros


#Spatial Analysis Relevant Data: 
-------------------------------
**paidperstate_flood** = Total USD paid *per state* for any flood adaptation  

**exsitu_paidperstate** = Total USD paid *per state* for properties bought out due to flooding
**insitu_paidperstate** = Total USD paid *per state* for properties to stay and adapt in place (in situ)

**floodproperties_bystate** = Total USD paid *per state* for all flooding-affected properties
**floodproperties_bystateperyear** = Total USD paid *per state, per year* for all flooding-affected properties

**exsitu_propsperstate** = Total USD paid *per state* for buyouts due to flooding 
**insitu_propsperstate** = Total USD paid *per state* for properties to adapt in place 

**propertiesbystate_allflood** = Joined all flood properties (in and ex situ) *by state* including total paid and number of properties 

**all_data_bystate** = Joined all datasets *by state* including all/insitu/exsitu for total amount paid and number of properties 

```{r setup, include=FALSE}
library(dplyr)
library(tidyverse)

library(tidyverse)
library(lubridate)
library(trend)
library(zoo)
library(Kendall)
library(tseries)

hazard_properties <- read.csv("./Data/Raw_Data/HazardMitigationAssistanceMitigatedProperties.csv")
```

#Note: 
I wrangled the data in two ways. First, like we talked about, I summarized amount paid over time, then looked at the amount paid for in-situ vs. ex-situ adaptation. However, all the observations before 1999 are NAs for amount paid and there are a few negative numbers (which I removed). Because of all this, I also summarized by number of properties in case we wanted to also look at that because we had data further back in time. 


<<<<<<< HEAD
#Actual Amount Paid by Year
```{r amount paid per year}
#All flood properties
colnames(hazardprops)
#Some negative values --> not sure what these were, so removed
clean_hazardprops <- hazard_properties %>% 
  filter(actualAmountPaid > 0)
```

#Actual Amount Paid 
```{r}
#All flood properties
colnames(hazard_properties)
clean_hazardprops <- hazard_properties %>%
  filter(!is.na(actualAmountPaid))%>%
  filter(actualAmountPaid > 0)

floodproperties_amountpaid <- clean_hazardprops %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10|202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(programFy)%>%
  summarise(Floodproperties_amountpaid = sum(actualAmountPaid))

#Ex situ
buyoutfloodproperties_amountpaid <- clean_hazardprops %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10', type))%>%
  group_by(programFy)%>%
  summarise(Buyouts_propertiesamountpaid = sum(actualAmountPaid))

#In situ
insitufloodproperties_amountpaid <- clean_hazardprops %>%
  filter(grepl('202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(programFy)%>%
  summarise(insituproperties_amountpaid = sum(actualAmountPaid))
```

```{r inflation adjustments}
install.packages("priceR")
library(priceR)
price <- floodproperties_amountpaid

amountpaid_inflation <- priceR::adjust_for_inflation(floodproperties_amountpaid$Floodproperties_amountpaid, floodproperties_amountpaid$programFy, "US", to_date = 2020)

inflation_floodproperties_amountpaid <- floodproperties_amountpaid %>%
  mutate("InflationAdjusted_AllAmountpaid" = amountpaid_inflation)

#Ex situ

buyouts_inflation <- priceR::adjust_for_inflation(buyoutfloodproperties_amountpaid$Buyouts_propertiesamountpaid, buyoutfloodproperties_amountpaid$programFy, "US", to_date = 2020)

inflation_buyoutfloodproperties_amountpaid <- buyoutfloodproperties_amountpaid%>%
  mutate("InflationAdjusted_BUYOUTAmountpaid" = buyouts_inflation)

#In situ
insitu_inflation <- priceR::adjust_for_inflation(insitufloodproperties_amountpaid$insituproperties_amountpaid, insitufloodproperties_amountpaid$programFy, "US", to_date = 2020)

inflation_insituprops_amountpaid <- insitufloodproperties_amountpaid%>%
  mutate("InflationAdjusted_INSITUAmountpaid" = insitu_inflation)

```


```{r visualize amount paid}

ggplot(floodproperties_amountpaid, aes(y= Floodproperties_amountpaid, x = programFy)) + 
  geom_line() +
  geom_smooth(method = lm)+
  labs(title = "All Flood Properties Amount Paid over Time",
       x= "Fiscal Year")

ggplot(inflation_floodproperties_amountpaid, aes(y= InflationAdjusted_AllAmountpaid, x = programFy)) + 
  geom_line() +
  geom_smooth(method = lm)+
  labs(title = "All Flood Properties Amount Paid over Time",
       x= "Fiscal Year")

##Ex-Situ Trends, and Inflation Adjusted Trends
ggplot(buyoutfloodproperties_amountpaid, aes(y= Buyouts_propertiesamountpaid, x = programFy)) + 
  geom_line() +
  geom_smooth(method = lm, aes(y= Buyouts_propertiesamountpaid, x = programFy), color = "coral", show.legend = FALSE)+
  labs(title = "Ex-Situ Flood Properties Amount Paid over Time")

ggplot(inflation_buyoutfloodproperties_amountpaid, aes(y= InflationAdjusted_BUYOUTAmountpaid, x = programFy)) + 
  geom_line() +
  geom_smooth(method = lm)+
  labs(title = "Inflation Adjusted: Ex Situ Flood Properties Amount Paid over Time",
       x= "Fiscal Year")

##In-Situ Trends and Inflation Adjusted Trends
ggplot(insitufloodproperties_amountpaid, aes(y= insituproperties_amountpaid, x = programFy)) + 
  geom_line() +
  geom_smooth(method = lm, aes(y= insituproperties_amountpaid, x = programFy), color = "coral", show.legend = FALSE)+
  labs(title = "In-Situ Flood Properties Amount Paid over Time")

ggplot(inflation_insituprops_amountpaid, aes(y= InflationAdjusted_INSITUAmountpaid, x = programFy)) + 
  geom_line() +
  geom_smooth(method = lm, aes(y= InflationAdjusted_INSITUAmountpaid, x = programFy), color = "coral", show.legend = FALSE)+
  labs(title = "In-Situ Flood Properties Amount Paid over Time")
 
```

#Amount paid per state
```{r Flood paid by state}

paidperstate_flood <- clean_hazardprops %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10|202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(state)%>%
  summarise(Floodproperties_amountpaid = sum(actualAmountPaid))%>%
  rename("STATE_NAME" = state)

#Ex situ
exsitu_paidperstate <- clean_hazardprops %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10', type))%>%
  group_by(state)%>%
  summarise(Buyouts_propertiesamountpaid = sum(actualAmountPaid))%>%
  rename("STATE_NAME" = state)

#In Situ
insitu_paidperstate <- clean_hazardprops %>%
  filter(grepl('202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(state)%>%
  summarise(insituproperties_amountpaid = sum(actualAmountPaid))%>%
  rename("STATE_NAME" = state)
```

Based on the graph and data for in-situ adaptation, 2008 looks like a bit of an outlier with a larger amount paid than the rest of the years, so I wanted to see what it looked like without 2008: 
```{r remove 2008 outlier}

no2008_insitufloodproperties <- insitufloodproperties_amountpaid%>%
  filter(programFy != 2008) 

ggplot(no2008_insitufloodproperties, aes(y= insituproperties_amountpaid, x = programFy)) + 
  geom_line() +
  geom_smooth(method = lm, aes(y= insituproperties_amountpaid, x = programFy), color = "coral", show.legend = FALSE)+
  labs(title = "In-Situ Flood Properties Amount Paid over Time")

```

#Number of Properties by Year
```{r Number of Properties by year}
#All flood properties
floodproperties <- hazard_properties %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10|202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(programFy)%>%
  summarise(Floodpropertiesperyear = sum(numberOfProperties))


Flood_properties_ts <- ts(floodproperties$Floodpropertiesperyear,
                   start=c(1989,1), frequency = 12) 
plot(Flood_properties_ts)

#Ex situ
buyoutfloodproperties <- hazard_properties %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10', type))%>%
  group_by(programFy)%>%
  summarise(Buyouts_propertiesperyear = sum(numberOfProperties))

buyouts_ts <- ts(buyoutfloodproperties$Buyouts_propertiesperyear,
                   start=c(1989,1), frequency = 12) 
plot(buyouts_ts)


#In situ
insitufloodproperties <- hazard_properties  %>%
  filter(grepl('202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(programFy)%>%
  summarise(insitupropertiesperyear = sum(numberOfProperties))

finalprojecttheme <- theme_light(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
            legend.position = "right",  
            plot.title = element_text(size=14, hjust=0.5), 
            plot.subtitle = element_text(size=10, hjust=0.5))

ggplot(floodproperties, aes(y= Floodpropertiesperyear, x = programFy)) + 
  geom_line() +
  geom_smooth(method = lm, aes(y= Floodpropertiesperyear, x = programFy), color = "coral", show.legend = FALSE)+
  labs(title = "Number of Hazard Mitigation Grant Recipient Properties 1985 - 2020", 
       x = "Fiscal Year", 
       y= "Number of Properties per Year") +
  finalprojecttheme

ggplot(buyoutfloodproperties, aes(y= Buyouts_propertiesperyear, x = programFy)) + 
  geom_line() +
  geom_smooth(method = lm, aes(y= Buyouts_propertiesperyear, x = programFy), color = "coral", show.legend = FALSE)+
  labs(title = "Number of Properties that Received Funding for Ex-Situ Adaptation to Flooding", 
  subtitle = "(1985 - 2020)")

ggplot(insitufloodproperties, aes(y= insitupropertiesperyear, x = programFy)) + 
  geom_line() +
  geom_smooth(method = lm, aes(y= insitupropertiesperyear, x = programFy), color = "coral", show.legend = FALSE)+
  labs(title = "Number of Properties that Received Funding for In-Situ Adaptation to Flooding", 
  subtitle = "(1985 - 2020)")

```

```{r number of properties by state}
#All flood properties
floodproperties_bystateperyear <- hazardprops %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10|202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(state, programFy)%>%
  summarise(Floodpropertiesperyear = sum(numberOfProperties))%>%
  rename("STATE_NAME" = state)

floodproperties_bystate <- hazardprops %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10|202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(state)%>%
  summarise(Floodpropertiesperyear = sum(numberOfProperties))%>%
  rename("STATE_NAME" = state)


#Ex situ
exsitu_propsperstate <- hazardprops %>%
  filter(grepl('200.1|200.2|200.3|200.4|200.5|200.6|200.7|200.8|200.9|201.1|201.2|201.3|201.4|201.5|201.6|201.7|201.8|201.9|201.10', type))%>%
  group_by(state)%>%
  summarise(Buyouts_propertiesperyear = sum(numberOfProperties))%>%
  rename("STATE_NAME" = state)

#In situ
insitu_propsperstate <- hazardprops %>%
  filter(grepl('202.1|202.2|202.3|202.4|203.1|203.2|203.3|203.4|204.1|204.1|204.2|204.3|204.4|204.5|204.6|207.2', type))%>%
  group_by(state)%>%
  summarise(insitupropertiesperyear = sum(numberOfProperties))%>%
  rename("STATE_NAME" = state)
```

#I combined the data in a variety of different ways depending on what would be the most useful for the analysis. There is one for each of the three categories (all flood properties, just ex-situ, just in-situ) with both the properties and total amount paid. There is also a dataframe with all of this information in one including the number of properties and amount paid for all flood properties, for in-situ, and for ex-situ. 
```{r join by year datasets}
propertiesvalue_allflood <- left_join(floodproperties, floodproperties_amountpaid, by = "programFy")

insitu_paidprops<- left_join(insitufloodproperties, insitufloodproperties_amountpaid, by = "programFy")

exsitu_paidprops <- left_join (buyoutfloodproperties, buyoutfloodproperties_amountpaid, by = "programFy") 

joinexsitu_insitu <- left_join (exsitu_paidprops, insitu_paidprops, by = "programFy")

all_data_propvalue <- left_join(propertiesvalue_allflood, joinexsitu_insitu, by = "programFy")
```

```{r join by state datasets}
propertiesbystate_allflood <- left_join(floodproperties_bystate, paidperstate_flood, by = "STATE_NAME")

insitu_state<- left_join(insitu_paidperstate, insitu_propsperstate, by = "STATE_NAME")

exsitu_state <- left_join (exsitu_paidperstate, exsitu_propsperstate, by = "STATE_NAME") 

state_in.ex.situ <- left_join (exsitu_state, insitu_state, by = "STATE_NAME")

all_data_bystate <- left_join(state_in.ex.situ , propertiesbystate_allflood, by = "STATE_NAME")
```



#Make NAs Zero, and Date column which is January 1st for each year.  *Note this NAs as an assumption we made in the discussion*
```{r}
zero_NA_alldata <- all_data_propvalue%>%
  mutate("Date" = paste("1/1/", programFy))

zero_NA_alldata$Date <- gsub(" ", "", zero_NA_alldata$Date)

zero_NA_alldata$Date <- as.Date(zero_NA_alldata$Date, "%m/%d/%Y")


zero_NA_alldata[is.na(zero_NA_alldata)] <- 0

```

#Bind Inflation Adjusted
```{r}
inflation_exsituinsitu <- left_join(inflation_buyoutfloodproperties_amountpaid, inflation_insituprops_amountpaid, by = "programFy")

allinflationadjusted <- left_join(inflation_floodproperties_amountpaid, inflation_exsituinsitu, by = "programFy")
```

inflation_floodproperties_amountpaid inflation_buyoutfloodproperties_amountpaid inflation_insituprops_amountpaid
```{r}
write.csv(all_data_bystate,"./Data/alldatabystate.csv")
write.csv(zero_NA_alldata, "./Data/alldatabyyear.csv")
write.csv(floodproperties_bystateperyear, "./Data/bystateandyearflood.csv")
write.csv(allinflationadjusted, "./Data/inflationadjusted.csv")
```


